If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-14 GroupPolarizationStatmod/paper/ms.tex
changed all betas to lambdas

To calculate the false discovery rate for each experimental condition, we first must
identify a constant latent mean, $\mu$, and latent variances $\sigmapre$ and
$\sigmapost$ that generate observation distributions\mt{define earlier?} with
reported mean pre- and post-deliberation opinions and resulting shift $s$ (calculated
using Equation~\ref{eq:group-polarization}). Define the collection of these
parameters $\beta = \{\mu, \sigmapre, \sigmapost\}$. We identify $\beta$ through a
hillclimbing optimization algorithm that finds $\beta$ that minimizes the difference
between reported pre- and post-deliberation mean opinions and simulated ones
calculated with Equation~\ref{eq:meanobs}.

The algorithm to find $\beta$ holds $\mu$ constant while searching for $\sigmapre$
and $\sigmapost$ that generate observed $\meanobspre$ and $\meanobspost$, within a
tolerance of $\pm10^{-3}$ (Algorithm~\ref{alg:hillclimbing}). 
The algorithm requires a researcher to specify an initial
guess for $\beta$ and auxiliary parameter settings, which generally vary between
experimental conditions.  Furthermore, the algorithm does not always produce
reasonable distributions with a given $\beta$, and sometimes it does not produce
spurious group polarization.  Therefore, it is necessary to sanity check the latent
and ordinal opinion distributions for the $\beta$ found by the algorithm. If all is
well, the experimental condition can be marked as plausibly spurious.  To tame the
complexity involved in data entry and model fitting for each of these sixty
experimental conditions, we developed a web application to input data from control
the algorithm and inspect its results (Figure~\ref{fig:webapp}). Web application code is
available online in our GitHub
repository\footnote{\url{https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/app.R}}.

@@

Note that so far, we have only identified cases where spurious group polarization plausibly
occurs in the asymptotic case where $N\to\infty$ ($N$ being the number of participants).
Experimental conditions yielding plausibly spurious group polarization warrant 
some suspicion. If there exists set of latent simple consensus 
parameters $\beta$ that create binned (Equation~\ref{eq:ordinal-frequency})
opinion distributions that naively seem to be group polarized, then the means
of at least some samples from these distibutions with empirical numbers of 
participants would also appear to be shifted consistent with group
polarization. This process itself does not tell us the false detection rate.
We use the identified simple consensus parameters to define pre- and post-deliberation
latent opinion distributions in simulated trials of group polarization 
experiments.

@@

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\beta_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
  \label{eq:cohens}
\end{equation}
\noindent
The family-wise error rate is, $\alpha_{e} = \Pr(D|\neg E)$, is the 
fraction of trials, 
with $d_{ei} > d^*$ for the $i^\text{th}$ trial for experimental
condition $e$, written
\begin{equation}
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
  \label{eq:study_aggregate}
\end{equation}
\noindent
where
\[
  \Theta(d_{ei}; d^*) = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
\]
and $N_T = 1000$ is the number of trials. For simplicity we assume that the sign of
$d_{ei}$ is positive, meaning a shift to greater extremisim. It is possible that
$d_{ei}$ could be negative, indicating a shift to lesser extremity. Indeed we
observe many simulations with a shift to lesser extremity. For now we do not
include these as indicating a ``significant'' effect, which we return to in the
Analysis below.

@@

We used our formal distribution model of group polarization experiments
to identify plausible latent simple consensus distributions that generate spurious
group polarization in observed ordinal distributions 
in the asymptotic case of taking $N \to \infty$ ordinal measurements.
Across the ten studies we analyzed, we excluded 3 out of 60 experimental
conditions that did not claim to be positive findings of group polarization.
Of the remaining 57, we identified simple consensus $\beta$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\beta$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. We rejected $\beta$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\beta$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices.

>>>>

To calculate the false discovery rate for each experimental condition, we first must
identify a constant latent mean, $\mu$, and latent variances $\sigmapre$ and
$\sigmapost$ that generate observation distributions\mt{define earlier?} with
reported mean pre- and post-deliberation opinions and resulting shift $s$ (calculated
using Equation~\ref{eq:group-polarization}). Define the collection of these
parameters $\lambda = \{\mu, \sigmapre, \sigmapost\}$. We identify $\lambda$ through a
hillclimbing optimization algorithm that finds $\lambda$ that minimizes the difference
between reported pre- and post-deliberation mean opinions and simulated ones
calculated with Equation~\ref{eq:meanobs}.

The algorithm to find $\lambda$ holds $\mu$ constant while searching for $\sigmapre$
and $\sigmapost$ that generate observed $\meanobspre$ and $\meanobspost$, within a
tolerance of $\pm10^{-3}$ (Algorithm~\ref{alg:hillclimbing}). 
The algorithm requires a researcher to specify an initial
guess for $\lambda$ and auxiliary parameter settings, which generally vary between
experimental conditions.  Furthermore, the algorithm does not always produce
reasonable distributions with a given $\lambda$, and sometimes it does not produce
spurious group polarization.  Therefore, it is necessary to sanity check the latent
and ordinal opinion distributions for the $\lambda$ found by the algorithm. If all is
well, the experimental condition can be marked as plausibly spurious.  To tame the
complexity involved in data entry and model fitting for each of these sixty
experimental conditions, we developed a web application to input data from control
the algorithm and inspect its results (Figure~\ref{fig:webapp}). Web application code is
available online in our GitHub
repository\footnote{\url{https://github.com/mt-digital/gp-statmod/blob/main/GroupPolarizationStatmod/app.R}}.

@@

Note that so far, we have only identified cases where spurious group polarization plausibly
occurs in the asymptotic case where $N\to\infty$ ($N$ being the number of participants).
Experimental conditions yielding plausibly spurious group polarization warrant 
some suspicion. If there exists set of latent simple consensus 
parameters $\lambda$ that create binned (Equation~\ref{eq:ordinal-frequency})
opinion distributions that naively seem to be group polarized, then the means
of at least some samples from these distibutions with empirical numbers of 
participants would also appear to be shifted consistent with group
polarization. This process itself does not tell us the false detection rate.
We use the identified simple consensus parameters to define pre- and post-deliberation
latent opinion distributions in simulated trials of group polarization 
experiments.

@@

To calculate the false detection rate we begin by simulating 
pre- and post-deliberation 
opinions by drawing $N_e$ samples from normal latent opinion
distributions that generated spurious group polarization, i.e., with
parameters labelled $\lambda_e$ found in the hillclimbing step. Binning these
random values simulates observations. $N_e$ is the number
of participants in experimental condition $e$. Each trial $i$ of $N_e$
simulated pre- and post-deliberation observations is then inspected to see
if spurious group polarization occurred in the simulation. Spurious group
polarization is determined by first inferring the latent parameters of each
simulated experimental trial by fitting an ordered probit model~\cite{Liddell2018}. 
The inferred parameters are then used to calculate the effect size for that
trial, Cohen's $d$, defined as~\cite[p. 331]{Liddell2018}
\begin{equation}
  d = \frac{\mupost - \mupre}{\sqrt{\frac{1}{2}(\sigmapost^2 + \sigmapre^2)}}.
  \label{eq:cohens}
\end{equation}
\noindent
The family-wise error rate is, $\alpha_{e} = \Pr(D|\neg E)$, is the 
fraction of trials, 
with $d_{ei} > d^*$ for the $i^\text{th}$ trial for experimental
condition $e$, written
\begin{equation}
  \alpha_e = \frac{1}{N_T} \sum_i^{N_T} \Theta(d_{ei}; d^*),
  \label{eq:study_aggregate}
\end{equation}
\noindent
where
\[
  \Theta(d_{ei}; d^*) = 
    \begin{cases}
      1 & \text{ if } d_{ei} > d^* \\
      0 & \text{ otherwise.}
    \end{cases}
\]
and $N_T = 1000$ is the number of trials. For simplicity we assume that the sign of
$d_{ei}$ is positive, meaning a shift to greater extremisim. It is possible that
$d_{ei}$ could be negative, indicating a shift to lesser extremity. Indeed we
observe many simulations with a shift to lesser extremity. For now we do not
include these as indicating a ``significant'' effect, which we return to in the
Analysis below.

@@

We used our formal distribution model of group polarization experiments
to identify plausible latent simple consensus distributions that generate spurious
group polarization in observed ordinal distributions 
in the asymptotic case of taking $N \to \infty$ ordinal measurements.
Across the ten studies we analyzed, we excluded 3 out of 60 experimental
conditions that did not claim to be positive findings of group polarization.
Of the remaining 57, we identified simple consensus $\lambda$ that generate spurious
group polarization in 54 experimental conditions. In the other three, we did
identify $\lambda$ that generate spurious group polarization, but the initial
simulated distributions were highly polarized. We rejected $\lambda$ generated for
two spurious group polarization from Schkade, et al., (2010) because
the histogram plots of the observed opinion distributions are not
polarized~\cite[Figure 1, p. 234]{Schkade2010}. The other condition where we
found but rejected spurious group polarization $\lambda$ was in a question 
about a ``bad teacher'' in Myers (1975)~\cite{Myers1975a}, where lower values
signify increasingly worse impressions of the teacher and higher values signify
increasingly good impressions. It is unlikely that a
``bad teacher'' would be rated as good in this case, so we excluded this 
from our analysis as well. This exclusion may be overly generous since Myers (1975)
did not provide details beyond mean values and opinion shifts, and the article was
published before open data practices.
